Ornaments as indicators of social change before and after European contact at Kiwulan, Northeastern Taiwan

### 2019-08-17 analysis/paper/paper.Rmd
revised abstract and added highlights, used viridis color for plots

---
title: "Ornaments as indicators of social change before and after European contact at Kiwulan, Northeastern Taiwan"
author:
  - Li-Ying Wang
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  European expansion to Asia since 16th century made a great impact on local indigenous people, where indigenous societies faced forced contact with imperial powers that often transformed their economic, cultural, and socio-political systems. In East Asia, direct European colonial rule on local communities was much less pronounced than it in island Southeast Asia and Oceania. However, the question of long-lasting indirect impacts on indigenous peoples in the periphery of colonial control remains largely unanswered. Here I use northeastern Taiwan as an example to explore the indirect European colonial impacts on local societies by examining trade ornaments excavated at Kiwualn site (1400-1900 AD), a large late Iron Age settlement that encountered different episodes of culture contact. The cultural context, variety, and spatial distribution of trade ornaments were examined in terms of three time episodes, before European contact, European contact, and Chinese contact. The results show that there is a greater amount and diversity of ornament types during the European contact period, and the spatial pattern of ornaments indicates non-random clusters that hints uneven distribution. Those ornaments may carry an exotic and powerful image that signals wealth and trading connections with Europeans for the inhabitants of Kiwulan. These symbolic values may have stimulated more competition between aggrandizing individuals for prestige and wealth accumulation resulting in an increase in social inequality. This study demonstrates the use of foreign ornaments as a proxy to detect indirect colonial influence on local indigenous populations and inform on social changes relating to trade with colonial groups. Ornaments can give insights into the amplification of social inequality stimulated by European colonization. It also shows the agency of indigenous people to incorporate ornaments into their social system.
keywords: |
  Historical archaeology; European colonization; Trade ornaments; Exchange networks; Taiwan; Monte Carlo simulation
highlights: |
  These are the highlights. 
---

@@

Prior to contact with the Europeans in the early 17th century, Taiwan had been involved in both regional and long-distance East and Southeast Asian trade systems since the tenth century, and encountered large waves of Han Chinese immigration after 17th century when Taiwan was incorporated into the territories of Qing Dynasty of China [@Liu2017; @Andrade2007, ch. 11]. The exchange network in northern Taiwan was influenced by long-distance trade between Fuzhou or Quanzhou in China and Ryukyu in Japan since the 15th century due to its location on shipping routes [@Chen2005]. This intensified in the 17th century with the arrival of the Europeans and their trade goods [@Wang2007].

@@

```{r, KWL-sam-area, fig.cap = "Map showing the largest section of excavation areas at Kiwulan, and the distribution of forty squares sampled in this paper presented in red with square ID number. Each square is 4 x 4 m", fig.width=image_width_2_col_inch}
# add to the map and scale and north arrow. Check it out after knitting 
library(sf)
library(cartography)
#devtools::install_github('oswaldosantos/ggsn')
library(ggsn)
# read in the spatial data
burial <- invisible(st_read(here("analysis", "data", "raw_data", "AD_burial.shp"), quiet = TRUE))
post <- invisible(st_read(here("analysis", "data", "raw_data", "AD_postholes.shp"), quiet = TRUE))
AD <- invisible(st_read(here("analysis", "data", "raw_data", "AD_zone.shp"), quiet = TRUE))
AD_data <- invisible(st_read(here("analysis", "data", "raw_data", "AD_withdata.shp"), quiet = TRUE))
location <- invisible(st_read(here("analysis", "data", "raw_data", "location.shp"), quiet = TRUE))
units_40 <- invisible(read.csv(here("analysis", "data", "raw_data", "kwl-list-of-sampling-squares.csv")))

@@

# mapping sampling area
KWL_excavation_map <- 
ggplot(sample_units) +
  geom_sf(data = AD_data, fill = NA) +
  geom_sf(color = 'red') +
  geom_sf_text(aes(label = `坑號`), size = 2.5, 
               alpha = 0.9, fontface = "bold") +
  geom_sf(data = post, color = 'blue', size = 0.2, alpha = 0.5) +
  theme_minimal() +
  blank() +
    north(AD_data, scale = 0.2, symbol = 3)
  #theme_void()

#1. Need to say that this is a map of excavation areas at Kiwulan, because the reader cannot see the site name anywhere in this figure.
#2. Need to add in here that each square is 4 x 4 m
#3. and say which way north is. Or add to the map and scale and north arrow.

@@

Previous researchers divided the Upper Layer Culture component into six layers spanning from the 14th century to the 19th century with hundred year intervals according to the radiocarbon dates, excavation depth, and types of diagnostic porcelains such as blue and white porcelains [@Hsieh2009; @Wang2011]. The layers were assigned numbers from 1 to 6 where L1 corresponds to the youngest layer and L6, the bottom layer, refers to the oldest layer. However, because the specific methods of assigning layers to a time period was not documented in detail for the previous chronology, I reexamined the excavation records to produce a new chronology for the sampled squares. For each sampled square I separated excavation units into three episodes of time: pre-European contact period (L5 and L6), European contact period (L4), and Chinese contact period (L2 and L1).

@@

```{r plot-all-spatial-pattern, fig.cap= "Spatial pattern of all class of ornament by time periods"}
library(sf)

# read in the spatial data
AD_data <- invisible(st_read(here("analysis", "data", "raw_data", "AD_withdata.shp"), quiet = TRUE))
units_40 <- invisible(read.csv(here("analysis", "data", "raw_data", "kwl-list-of-sampling-squares.csv")))
# tidy spatial data for 40 sampling units
sample_units <-
  AD_data %>%
  semi_join(units_40, by = c (`坑號` = "the_sq"))
# join two dataset to get coordinate

ornaments_period_join <-
  ornaments_period %>%
  mutate(period = as.factor(period)) %>% # seems need this line first
  group_by(Pit_No, period) %>%
  count(Categories) %>%
  inner_join(AD_data, by = c("Pit_No" = "坑號"))
# plot spatial pattern for all ornaments

plot_spatial_distribution_all_ornaments_by_period <- 
ggplot() +
  geom_sf(data = sample_units, fill = NA) +
  geom_sf(data = ornaments_period_join, 
          aes(geometry = geometry, 
              fill =n), 
          alpha = 0.9) +
  facet_wrap(~ period) +
  labs(fill = "frequency") +
  scale_fill_gradient(high = "black", low = "grey") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        strip.text.x = element_text(size = 11))

@@

plot_spatial_distribution_some_ornaments_by_period <- 
ggplot() +
  geom_sf(data = sample_units, fill = NA) +
  geom_sf(data = ornaments_period_join_selected_types, 
          aes(geometry = geometry,
              fill = n), alpha = 0.9) +
  facet_grid(rows = vars(Categories),
             cols = vars(period)) +
  labs(title = "",
       fill = "frequency") +
  scale_fill_gradient(high = "black", low = "grey") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        strip.text.x = element_text(size = 8))

@@

The results show that there is one major core area during the pre-European contact period, multiple core areas during European contact period, and a single core during the Chinese contact period (Figure \@ref(fig:plot-all-densities)). Those maps present three consistent sub-regions with a core area that shifts over time. The distribution might indicate different social groups who possessed more ornaments. The multiple groups during European contact period might reflect more unequal consumption of ornaments across the site, relative to other periods. However, the generation of core areas might be biased due to small sample size, for example, a few ornaments found at one single square during the Chinese period could create an obvious hot spot. To further evaluate the pattern, we test the hypothesis that the distribution of ornaments is not randomly distributed to determine if social processes might explain ornament distributions.
```{r function for ANN values}
# function for hypothesis testing
nndist_hypo_testing <- function(x, n = 1e3L,  ... ) {

@@

# Acknowledgements

<!-- The following line inserts a page break when the output is MS Word. For page breaks in PDF, use \newpage on its own line.  -->
##### pagebreak

>>>>

---
title: "Ornaments as indicators of social change before and after European contact at Kiwulan, Northeastern Taiwan"
author:
  - Li-Ying Wang
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  The direct impact of European colonization since 16th century on local indigenous people has been widely discussed. In East Asia, direct European colonial rule on local communities was much less pronounced. However, the question of long-lasting indirect impacts on indigenous peoples in the periphery of colonial control remains largely unanswered. Using trade ornaments recovered from Kiwualn site (1400-1900 AD) in northeastern Taiwan, this research explores the indirect European colonial impacts on local societies by examining cultural context, variety, and spatial distribution of trade ornaments in terms of three time episodes relating different scales of culture contact. The results indicate the diversity of ornaments was greater and their spatial distribution was non-random during the European contact period that hints the increasing social inequality in a colonial context. This study demonstrates the use of foreign ornaments as a proxy to detect indirect colonial influence on local indigenous populations and inform on social changes relating to trade with colonial groups. Ornaments can give insights into the amplification of social inequality stimulated by European colonization. It also shows the agency of indigenous people to incorporate ornaments into their social system.
keywords: |
  Historical archaeology; European colonization; Trade ornaments; Exchange networks; Taiwan; Monte Carlo simulation
highlights: |
  1. Indirect impact of European colonization on indigenous peoples in the periphery of colonial control was rarely discussed.
  2. Evidence of the quantity, variety, and spatial distribution of trade ornaments in both local cultural and colonial context is provided.
  3. Results of hypothesis testing point to greater amount and diversity of ornament types with non-random, and most highly concentrated spatial clusters during the European contact.
  4. Ornaments may carry an exotic and powerful image that signals wealth and trading connections with colonists that resulted in an increase in social inequality due to competition of wealth accumulation between individuals.
  5. Ornaments can be a good proxy to detect indirect colonial influence on local indigenous populations and inform on social changes relating to trade with colonial groups.
---

@@

Prior to contact with the Europeans in the early 17th century, Taiwan had been involved in both regional and long-distance East and Southeast Asian trade systems since the 10th century, and encountered large waves of Han Chinese immigration after 17th century when Taiwan was incorporated into the territories of Qing Dynasty of China [@Liu2017; @Andrade2007, ch. 11]. The exchange network in northern Taiwan was influenced by long-distance trade between Fuzhou or Quanzhou in China and Ryukyu in Japan since the 15th century due to its location on shipping routes [@Chen2005]. This intensified in the 17th century with the arrival of the Europeans and their trade goods [@Wang2007].

@@

```{r, KWL-sam-area, fig.cap = "Map showing the largest section of excavation areas at Kiwulan, and the distribution of forty squares sampled in this paper presented in red with square ID number. Small dots represent the location of post-holes. Each square is 4 x 4 m", fig.width=image_width_2_col_inch}
# add to the map and scale and north arrow. Check it out after knitting 
library(sf)
library(cartography)
#devtools::install_github('oswaldosantos/ggsn')
library(ggsn)
# read in the spatial data
burial <- invisible(st_read(here("analysis", "data", "raw_data", "AD_burial.shp"), quiet = TRUE))
post <- invisible(st_read(here("analysis", "data", "raw_data", "AD_postholes.shp"), quiet = TRUE))
AD <- invisible(st_read(here("analysis", "data", "raw_data", "AD_zone.shp"), quiet = TRUE))
AD_data <- invisible(st_read(here("analysis", "data", "raw_data", "AD_withdata.shp"), quiet = TRUE))
location <- invisible(st_read(here("analysis", "data", "raw_data", "location.shp"), quiet = TRUE))
units_40 <- invisible(read.csv(here("analysis", "data", "raw_data", "kwl-list-of-sampling-squares.csv")))

@@

# mapping sampling area
KWL_excavation_map <- 
ggplot(sample_units) +
  geom_sf(data = AD_data, fill = NA) +
  geom_sf(color = 'red') +
  geom_sf_text(aes(label = `坑號`), size = 2.5, 
               alpha = 0.9, fontface = "bold") +
  geom_sf(data = post, color = 'blue', size = 0.2, alpha = 0.4) +
  theme_minimal() +
  blank() +
    north(AD_data, scale = 0.2, symbol = 3)
  #theme_void()

@@

Previous researchers divided the Upper Layer Culture component into six layers spanning from the 14th century to the 19th century with hundred year intervals according to the radiocarbon dates, excavation depth, consistency of contexts, and types of diagnostic porcelains such as blue and white porcelains [@Hsieh2009; @Wang2011]. The layers were assigned numbers from 1 to 6 where L1 corresponds to the youngest layer and L6, the bottom layer, refers to the oldest layer. However, because the specific methods of assigning layers to a time period was not documented in detail for the previous chronology, I reexamined the excavation records to produce a new chronology for the sampled squares. For each sampled square I separated excavation units into three episodes of time: pre-European contact period (L5 and L6), European contact period (L4), and Chinese contact period (L2 and L1).

@@

```{r plot-all-spatial-pattern, fig.cap= "Spatial pattern of all class of ornament by time periods"}
library(sf)
library(viridis)

# read in the spatial data
AD_data <- invisible(st_read(here("analysis", "data", "raw_data", "AD_withdata.shp"), quiet = TRUE))
units_40 <- invisible(read.csv(here("analysis", "data", "raw_data", "kwl-list-of-sampling-squares.csv")))

# tidy spatial data for 40 sampling units
sample_units <-
  AD_data %>%
  semi_join(units_40, by = c (`坑號` = "the_sq"))

# plot spatial pattern for all ornaments
plot_spatial_distribution_all_ornaments_by_period <- 
ggplot() +
  geom_sf(data = sample_units, fill = NA) +
  geom_sf(data = ornaments_period_join, 
          aes(geometry = geometry, 
              fill =n), 
          alpha = 0.9) +
  facet_wrap(~ period) +
  labs(fill = "frequency") +
  scale_fill_viridis(direction = -1) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        strip.text.x = element_text(size = 11))

@@

plot_spatial_distribution_some_ornaments_by_period <-
ggplot() +
  geom_sf(data = sample_units, fill = NA) +
  geom_sf(data = ornaments_period_join_selected_types, 
          aes(geometry = geometry,
              fill = n), alpha = 0.9) +
  facet_grid(rows = vars(Categories),
             cols = vars(period)) +
  labs(title = "",
       fill = "frequency") +
  scale_fill_viridis(direction = -1) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        strip.text.x = element_text(size = 8))

@@

The results show that there is one major core area during the pre-European contact period, multiple core areas during European contact period, and a single core during the Chinese contact period (Figure \@ref(fig:plot-all-densities)). Those maps present three consistent sub-regions with a core area that shifts over time. The distribution might indicate different social groups who possessed more ornaments. The multiple groups during European contact period might reflect more unequal consumption of ornaments across the site, relative to other periods. However, the generation of core areas might be biased due to small sample size, for example, a few ornaments found at one single square during the Chinese period could create an obvious hot spot. To further evaluate the pattern, we test the hypothesis that the distribution of ornaments is not randomly distributed to determine if social processes might explain ornament distributions.

```{r function for ANN values}
# function for hypothesis testing
nndist_hypo_testing <- function(x, n = 1e3L,  ... ) {

@@

# Acknowledgements
<!-- The following line inserts a page break when the output is MS Word. For page breaks in PDF, use \newpage on its own line.  -->
